```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```
---
title: "<h1 style='color:#2C3E50; font-size:36px; text-align:center;'>
           Clustering the Diabetes Dataset with K-Means </h1>"
author: "<h4 style='text-align:center;'>Kawtar Zouhair</h4>"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    number_sections: yes
    toc: yes
    theme: cosmo
    highlight: tango
---


# **Introduction**

K-means is an unsupervised machine learning algorithm used to find groups of observations (clusters) that share similar characteristics. In this tutorial, we will use a [Diabetes Dataset](https://www.kaggle.com/akram24/diabetes-pca) to cluster different types of diabetes data. The dataset contains various health parameters related to diabetes.

# **Loading Data** {.tabset .tabset-fade .tabset-pills}

First, let's load the necessary libraries and read the data.


```{r}
# Load libraries
library(tidyverse)
library(corrplot)
library(gridExtra)
library(GGally)
library(knitr)
library(cluster)
library(factoextra)
```


# Load the diabetes dataset
```{r}
diabetes_raw=read.csv("C:/Users/kawta/Desktop/Module/Data Science/Diabetes/subfolder/diabetes_dataset.csv")

```

# Cleaning and Prepping the Data

```{r}
# Load necessary libraries
library(dplyr)

# Assign the raw data to a new variable
diabetes <- diabetes_raw

# Outcome is a categorical variable with values 0 or 1, convert it to a factor
diabetes$Outcome <- factor(diabetes$Outcome)

# Spell out the Outcome flag (1 = Diabetes, 0 = healthy)
diabetes$Outcome <- ifelse(diabetes$Outcome == "1", "Diabetes", "healthy")

# Rename the 'Outcome' column to 'target'
colnames(diabetes)[ncol(diabetes)] <- "target"

# Entfernen der 0-Werte aus der SkinThickness-Spalte
diabetes$SkinThickness[diabetes$SkinThickness == 0] <- NA

# Create two subsets: diabetes patients and healthy individuals
diabetes_patients <- subset(diabetes, target == "Diabetes")
diabetes_healthy <- subset(diabetes, target == "healthy")

# Ensure Insulin is numeric and replace 0 values with NA
diabetes$Insulin <- as.numeric(diabetes$Insulin)
diabetes$Insulin[diabetes$Insulin == 0] <- NA

# Cleaning the Blood Pressure Data: Replace 0 values with NA
diabetes$BloodPressure[diabetes$BloodPressure == 0] <- NA

# Check the number of missing values in the Blood Pressure column
sum(is.na(diabetes$BloodPressure))


#  replacing NA values in BloodPressure with the median (as a simple imputation strategy):
diabetes$BloodPressure <- ifelse(is.na(diabetes$BloodPressure), 
                                  median(diabetes$BloodPressure, na.rm = TRUE), 
                                  diabetes$BloodPressure)
# Entfernen der 0-Werte aus den Spalten Glucose und BMI
diabetes$Glucose[diabetes$Glucose == 0] <- NA
diabetes$BMI[diabetes$BMI == 0] <- NA

# Optional: Überprüfen der Anzahl der fehlenden Werte in Glucose und BMI
sum(is.na(diabetes$Glucose))
sum(is.na(diabetes$BMI))


```
# Show a preview of the data
```{r}
head(diabetes)
```
# Summary of the dataset
```{r}
summary(diabetes)
```
# Structure of the dataset
```{r}
str(diabetes)
```
# Histogram for each Attribute
```{r}
diabetes %>%
  gather(Attributes, value, 1:8) %>%
  ggplot(aes(x=value, fill=Attributes)) +
  geom_histogram(colour="black", show.legend=FALSE) +
  facet_wrap(~Attributes, scales="free_x") +
  labs(x="Values", y="Frequency",
       title="Diabetes Attributes - Histograms") +
  theme_bw()
```

# Boxplot for each Attribute
```{r}
diabetes %>%
  gather(Attributes, values, 1:8) %>%
  ggplot(aes(x=reorder(Attributes, values, FUN=median), y=values, fill=Attributes)) +
  geom_boxplot(show.legend=FALSE) +
  labs(title="Diabetes Attributes - Boxplots") +
  theme_bw() +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank()) +
  ylim(0, 35) +
  coord_flip()
```
# Correlation matrix 
```{r}
corrplot(cor(diabetes[,1:8]), type="upper", method="ellipse", tl.cex=0.9)
```
# Normalization
```{r}
diabetesNorm <- as.data.frame(scale(diabetes[,1:8]))
```
# Show the difference before and after normalization
```{r}
p1 <- ggplot(diabetes, aes(x=Glucose, y=BloodPressure)) +
  geom_point() +
  labs(title="Original data") +
  theme_bw()

p2 <- ggplot(diabetesNorm, aes(x=Glucose, y=BloodPressure)) +
  geom_point() +
  labs(title="Normalized data") +
  theme_bw()

```
# Subplot
```{r}
grid.arrange(p1, p2, ncol=2)
```

# Apply k-means with k=2
```{r}
# Replace NA values with the median for each specified column
diabetes$SkinThickness[is.na(diabetes$SkinThickness)] <- median(diabetes$SkinThickness, na.rm = TRUE)
diabetes$BMI[is.na(diabetes$BMI)] <- median(diabetes$BMI, na.rm = TRUE)
diabetes$Glucose[is.na(diabetes$Glucose)] <- median(diabetes$Glucose, na.rm = TRUE)
diabetes$Insulin[is.na(diabetes$Insulin)] <- median(diabetes$Insulin, na.rm = TRUE)
diabetes$BloodPressure[is.na(diabetes$BloodPressure)] <- median(diabetes$BloodPressure, na.rm = TRUE)


set.seed(1234)

# Ensure the data is a numeric matrix before running k-means
diabetes_k2 <- kmeans(as.data.frame(scale(diabetes[, 1:8])), centers = 2)

```

# View cluster centers
```{r}
diabetes_k2$centers
```
# Cluster assignments for each data point
```{r}
diabetes_k2$cluster
```
# Cluster sizes
```{r}
diabetes_k2$size
```
# Visualize the clusters
```{r}
table(diabetes_k2$cluster, diabetes$Outcome)
bss <- numeric()
wss <- numeric()
```
# Run k-means for different values of k
```{r}
set.seed(1234)

for(i in 1:10){
  bss[i] <- kmeans(diabetesNorm, centers=i)$betweenss
  wss[i] <- kmeans(diabetesNorm, centers=i)$tot.withinss
}
```

# Plot the between-cluster sum of squares vs. k
```{r}
p3 <- qplot(1:10, bss, geom=c("point", "line"), 
            xlab="Number of clusters", ylab="Between-cluster sum of squares") +
  scale_x_continuous(breaks=seq(0, 10, 1)) +
  theme_bw()

# Plot the total within-cluster sum of squares vs. k
p4 <- qplot(1:10, wss, geom=c("point", "line"),
            xlab="Number of clusters", ylab="Total within-cluster sum of squares") +
  scale_x_continuous(breaks=seq(0, 10, 1)) +
  theme_bw()
```
# Subplot
```{r}
grid.arrange(p3, p4, ncol=2)
```

# Apply k-means with k=3
```{r}
set.seed(1234)
diabetes_k3 <- kmeans(diabetesNorm, centers=3)
```
# View cluster assignments and customer segment comparison
```{r}
table(diabetes$Outcome, diabetes_k3$cluster)
```
# Cluster centers for k=3
```{r}
diabetes_k3$centers
```
# Boxplots for each cluster
```{r}
diabetes %>%
   mutate(k3_cluster = factor(diabetes_k3$cluster)) %>%
   gather(key = "variable", value = "value", -Outcome, -k3_cluster) %>%
   ggplot(aes(x=k3_cluster, y=value, colour=k3_cluster, groups=k3_cluster)) +
   facet_wrap(~variable, scales="free_y") + 
   geom_boxplot() + 
   theme_bw()
```
# Silhouette plot
```{r}
sil <- silhouette(diabetes_k3$cluster, daisy(diabetesNorm))
```

# Visualize the silhouette plot
```{r}
fviz_silhouette(sil, label=FALSE, print.summary=FALSE)
```

